---
title: "MIDAS package vignette"
output: 
  rmarkdown::html_vignette:
    number_sections: true
vignette: >
  %\VignetteIndexEntry{MIDAS_vignette}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE, 
  warning = FALSE,
  eval = FALSE)
```

```{r load package, echo=FALSE}
library(MIDAS)
library(MASS)
library(psych, quietly = TRUE)
library(scam, quietly = TRUE)
library(pracma, quietly = TRUE)
library(vegan, quietly = TRUE)
```

# Overview

The MIDAS package (v0.1.0) implements the **MI**crobiome **DA**ta **S**imulator functions that simulate realistic microbiome data, maintaining the distributional and taxon-taxon correlation of a template microbiome datase. Users may also change the parameter setup in the model to introduce ``effect'' to facilitate simulation.  

## Required packages 

Required packages for functions in MIDAS include: psych, MASS, vegan, pracma, scam. These packages are all available on CRAN.

## Installation

Install from the local file (the .tar.gz file), which you can download from https://github.com/mengyu-he/MIDAS.

```{r}
devtools::install_local("MIDAS_0.1.0.tar.gz", dependencies = TRUE)
```

A github installation is also available via

```{r}
install_github("mengyu-he/MIDAS")
```


## Open the Vignette in R

```{r}
browseVignettes("MIDAS")
```

or

```{r}
vignette("MIDAS_vignette", package = "MIDAS")
```


# MIDAS: Microbiome Data Simulator

There are three main functions in MIDAS package: *Midas.setup()*, *Midas.modify()*, *Midas.sim()*.

## Template data description

We demonstrate the method of MIDAS and functions in this package using a filtered microbiome dataset of patients with IBD(Inflammatory Bowel Disease) in Human Microbiome Project 2 (HMP2) [1]. A filtered version of the data is included in the MIDAS package and can be directly loaded through 

```{r}
data("count.ibd")
```

This "count.ibd" were a modified version of the IBD data in package **HMP2Data** on Bioconductor by filtering out samples with library sizes smaller than 3000 and taxa that appear in less than 2 samples. The filtered data have 146 rows (samples) and 614 columns (taxa).

```{r}
library(HMP2Data)
IBD16S()
```

## Template data description

We demonstrate the method of MIDAS and functions in this package using a filtered microbiome dataset of patients with IBD(Inflammatory Bowel Disease) in Human Microbiome Project 2 (HMP2) [1]. The data can be directly loaded through 

```{r}
data("count.ibd")
```

This HMP2 data were modified through package **HMP2Data** on Bioconductor by filtering out samples with library sizes smaller than 3000 and taxa that appear in less than 2 samples. The filtered data have 146 rows (samples) and 614 columns (taxa).

```{r, eval = FALSE}
library(HMP2Data)
IBD16S()

count.ibd <- t(IBD16S_mtx[, colSums(IBD16S_mtx)>3000] )
count.ibd <- count.ibd[, colSums(count.ibd>0)>1]

dim(count.ibd)
```


Other datasets come with the package include a vaginal microbiome dataset from HMP2 [1] and a upper-respiratory-tract microbiome dataset [2].

## Setup MIDAS model

The function *Midas.setup()* fits the underlying MIDAS model and extracts information from the template data and returns the estimated parameters for microbiome simulation.

```{r}
fitted <- Midas.setup(count.ibd, n.break.ties = 100, fit.beta = F)
```

The argument `n.break.ties` specifies he number of replicates used to break ties when ranking relative abundances for each taxon. When `fit.beta = FALSE` (default), <span style="color:red">a rank based approach is used to obtain the relative abundances for taxa that are considered to be non-zero in the simulated data. </span> When `fit.beta = TRUE`,  a beta distribution is fitted for the relative abundance of each taxon by matching the first two moments. Suppose for taxon $j$, the mean and sample variance of relative abundances $\boldsymbol{\pi}_j$ are $\bar{\boldsymbol{\pi}}_j$ and $s^2_j$, then the shape parameters of the beta distribution are
$$\alpha_j = \bar{\boldsymbol{\pi}}_j(\frac{\bar{\boldsymbol{\pi}}_j(1-\bar{\boldsymbol{\pi}}_j)}{s^2_j} -1 ),$$

$$\beta_j = \alpha_j\frac{1-\bar{\boldsymbol{\pi}}_j}{\bar{\boldsymbol{\pi}}_j}.$$
The returned values of *Midas.setup()* include: estimated tetrachoric correlation of observed presence-absence data (tetra.corr), estimated Pearson correlation of observed relative abundances (corr.rel.corrected), proportion of non-zero cells for each subject and taxon (pbar, eta), observed non-zero relative abundances of each taxon (non.zero.rel), observed mean relative abundances of each taxon (obs.rel.abund).

## Modify the fitted MIDAS model

The function *Midas.modify()* allows users to modify taxon relative abundances, library sizes, taxon proportion of non-zero cells and number of samples. Note that this is a **required** step even if no adjustment is made.

### No modification at all

When no additional adjustment is wanted (`lib.size = NULL`, `n.sample = NULL`, `rel.abund = NULL`, `taxa.1.prop = NULL`), the target library sizes of the simulated data are the same as that of the template. The following code is an example.

```{r}
fitted.modified <- Midas.modify(fitted, 
                               lib.size = NULL, 
                               rel.abund = NULL, 
                               taxa.1.prop = NULL)
```

### Modify library sizes

When users are interested in changing some key features of the microbiome community, we can use the following code.

```{r}
fitted.modified <- Midas.modify(fitted, 
                                lib.size = sample(1000:10000, 
                                                  nrow(count.ibd), 
                                                  replace = T) )
```

Here, the library sizes for all samples are taken as random samples with replacement from 1000 to 10000. The total number of samples is the same as in the IBD data. Note, if both library sizes and sample size are specified, the length of the vector of library sizes should be the same as n.sample. If the function gives the error that the marginal propotion of non-zeros can not be reached, 


### Modify features of taxa

Midas.modify() allows users to modify three types of quantities,

\item $\hat{p}_j$: mean relative abundances of taxa (`rel.abund`)
\item $\hat{p}^1_j$: mean relative abundances of taxa among non-zero samples (`rel.abund.1`)
\item $\hat{\delta}_j$: proportion of non-zero cells for taxa (`taxa.1.prop`)

These tree quantities are related by
$$\hat{p}_j = \hat{\delta}_j\hat{p}^1_j,$$
and the function will stop and give a message when the three quantities are all specified but the equation is not met. Other constraints include,

$$\hat{p}_j\leq \hat{p}^1_j$$
$$\hat{p}_j\leq \hat{\delta}_j$$

#### Only one feature of taxa is given

If only one of the above three quantities is given, Midas.modify() allows users to only provide information on mean relative abundances of taxa $\hat{p}$ or proportion of non-zero cells for taxa $\hat{\delta}$. Note that, only specifying the mean relative abundances of taxa among non-zero samples $\hat{p}^1$ is not allowed.

When only the mean relative abundances of taxa $p$ is given, the function fits a Shape Constrained Additive model (SCAM) model for $\delta$ on $p$,

$$log(\frac{\delta_j}{1-\delta_j}) = g(log10(p_{j})),$$
where $g$ is a monotone smoothing spline, $\delta_j$ is the observed proportion of non-zero cells of $j$-th taxon, and $p_{j}$ is the observed mean relative abundances of $j$-th taxon. 

The following is an example. We first generate a set of target taxa relative abundances by adding a normally distributed disturbance to the observed relative abundances. Let
$$\hat{p}_j = p_j + 0.1\cdot N(p_j, p_j),$$
where $N(p_j, p_j)$ is a normal distribution with mean and standard deviation equal to $p_j$. The following is the code for generating a set of target relative abundances as described, 

```{r}
n.taxa <- ncol(count.ibd)
target.rel.abund <- NULL

for (j in 1:n.taxa) {
  target.rel.abund[j] <- -1
  while (target.rel.abund[j]<0 | target.rel.abund[j]>1) {
    target.rel.abund[j] <- fitted$obs.rel.abund[j] + 0.1*(rnorm(1, fitted$obs.rel.abund[j], fitted$obs.rel.abund[j]))
  }
}
```

then the target taxa relative abundances can be used in Midas.modify() as the following,

```{r}
fitted.modified <- Midas.modify(fitted, 
                                rel.abund  = target.rel.abund)
```


When only the proportion of non-zero cells of taxa $\hat{\delta}$ is given, the function fits a SCAM model for $p$ on $\delta$,

$$log(\frac{p_j}{1-p_j}) = g(log10(\delta_{j})).$$
Suppose the target proportions of non-zero cells are the randomly shuffled ones from the template, the following code can be used,

```{r}
fitted.modified <- Midas.modify(fitted,
                                taxa.1.prop = sample(fitted$prop.1))
```


#### Full information of taxa features is given

If the three quantities $\hat{p}, \hat{p}^1,\hat{\delta}$, or any two of them are given, full information of taxa features is given, then no additional model for changing taxa features is needed. 

The following code is one example that gives the mean relative abundances $\hat{p}$ and uses the same proportion of non-zeros as in the template.

```{r}
fitted.modified <- Midas.modify(fitted,
                                rel.abund = target.rel.abund,
                                taxa.1.prop = "same")
```

Note that, specifying taxa features to be `"same"` is different from giving a `NULL` value. If in the above example, we use `taxa.1.prop = NULL`, then the proportion of non-zeros will be changed by fitting a SCAM model and making predictions; however, if we use `taxa.1.prop = "same"`, the proportions of non-zeros of taxa will remain the same as in the template data. 


#### Adjustment for parameters of beta distribution

If we set `fit.beta = TRUE` in Midas.setup(), which asks for simulating relative abundances from a beta distribution, Midas.modify() will make corresponding adjustment to the estimated shape parameters of the beta distribution. Suppose the observed relative abundances for $j$-th taxon are $\pi_{\cdot j}$, then the adjusted relative abundances are obtained by $\hat{\pi}_{\cdot j}= \pi_{\cdot j}^\alpha$, where $\alpha$ is chosen so that, after adjustment, the mean relative abundance in taxon j is $\hat{p}_{j}$. 


### Modify library sizes and features of taxa simultaneously

Library sizes and taxa features can be manipulated simultaneously. In the following code, the library sizes for all samples are taken as random samples with replacement from 1000 to 10000, and the zero proportions for all taxa were taken as a random permutation of the estimated zero proportions. 

```{r}
fitted.modified <- Midas.modify(fitted, 
                                lib.size = sample(1000:10000, nrow(count.ibd), replace = T),
                                taxa.1.prop  = sample(fitted$prop.1),
                                method.total0 = "scam")
```

Users are also able to modify library sizes and taxa relative abundances simultaneously through Midas.modify(),

```{r}
fitted.modified <- Midas.modify(fitted, 
                                lib.size = sample(1000:10000, nrow(count.ibd), replace = T),
                                rel.abund  = target.rel.abund,
                                method.total0 = "scam")
```

The function allows users to choose a method for finding a proper value of the total number of non-zeros that agrees with both target library sizes and proportion of non-zeros through the argument `method.total0`. By default, Midas.modify() uses `method.total0 = "scam"`, which fits a Shape Constrained Additive model (SCAM) model to estimate the presence-absence information for each taxon of each subject. If `method.total0 = "geometric"` is specified, the function uses the geometric mean of the number of total non-zeros cells predicted by target library sizes and taxa prevalence. The latter option is usually faster than the former one by avoiding fitting a model.


### Simulate microbiome data with the fitted MIDAS model

The function *Midas.sim()* takes the estimates from MIDAS model and simulates microbiome data. By default, the output includes the simulated presence-absence data, relative abundances and count data.The generated sample size will be the same as what is specified in the *Midas.modify* function. 

```{r}
simu <- Midas.sim(fitted.modified)
```


## References

[1] Lloyd-Price, J., Arze, C., Ananthakrishnan, A.N. *et al*. Multi-omics of the gut microbial ecosystem in inflammatory bowel diseases. *Nature* 569, 655–662 (2019). https://doi.org/10.1038/s41586-019-1237-9.

[2] Charlson, E.S., Chen, J., Custers-Allen, R. *et al*. Disordered microbial communities in the upper respiratory tract of cigarette smokers. *PloS one*, 5(12), e15216 (2010). https://doi.org/10.1371/journal.pone.0015216

